---
title: "Avoidable mortality: exploratory overview"
subtitle: "Deaths aged 0–74, preventable and treatable causes"
date: 2026-10-06
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: show
    fig-width: 8
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
library(tidyverse)
library(targets)
library(gtsummary)

theme_set(theme_minimal(base_size = 12))
pal_fill <- function(...) scale_fill_viridis_d(..., option = "D", end = 0.95)
pal_col  <- function(...) scale_colour_viridis_d(..., option = "D", end = 0.95)
```

This report describes `mort_count`, the output of
`preprocess_mortality()`.

It assumes **one row per death**, with the columns `sex`,
`area_residenza`, `asst`, `eta` (age in years), `group` (cause
group) and `cause` (specific cause). If your file is instead
pre-aggregated with a count column `n`, expand it once with
`mort_count <- tidyr::uncount(mort_count, n)` and everything below works
unchanged.

```{r}
#| label: data
tar_load(mort_count)
glimpse(mort_count)
```

# Univariate distributions

## Sex

```{r}
#| label: uni-sex
ggplot(mort_count, aes(x = sesso, fill = sesso)) +
  geom_bar() +
  pal_fill(guide = "none") +
  labs(title = "Deaths by sex", x = NULL, y = "Deaths")
```

## Area of residence

```{r}
#| label: uni-area
#| fig-height: 7
mort_count |>
  mutate(area_residenza = fct_lump_n(area_residenza, n = 25)) |>
  ggplot(aes(y = fct_rev(fct_infreq(area_residenza)))) +
  geom_bar(fill = "#3b528b") +
  labs(title = "Deaths by area of residence (top 25)", x = "Deaths", y = NULL)
```

## ASST (2022-2024)

```{r}
#| label: uni-asst
#| fig-height: 7
ggplot(mort_count, aes(y = fct_rev(fct_infreq(asst)))) +
  geom_bar(fill = "#3b528b") +
  labs(title = "Deaths by ASST (2022-2024)", x = "Deaths", y = NULL)
```

## Age at death

Note that the population is restricted to premature deaths (under 75),
so the distribution is bounded at 74.

```{r}
#| label: uni-eta
ggplot(mort_count, aes(x = eta)) +
  geom_histogram(binwidth = 5, boundary = 0,
                 fill = "#3b528b", colour = "white") +
  labs(title = "Age-at-death distribution", x = "Age (years)", y = "Deaths")
```

## Cause group

```{r}
#| label: uni-group
#| fig-height: 6
ggplot(mort_count, aes(y = fct_rev(fct_infreq(group)))) +
  geom_bar(fill = "#3b528b") +
  labs(title = "Deaths by cause group", x = "Deaths", y = NULL)
```

## Mechanism

`mechanism` records the avoidability lever each death is attributed to
(the preventable mechanisms together with the treatable arms).

```{r}
#| label: uni-mechanism
#| fig-height: 5
ggplot(mort_count, aes(y = fct_rev(fct_infreq(mechanism)))) +
  geom_bar(fill = "#3b528b") +
  labs(title = "Deaths by mechanism", x = "Deaths", y = NULL)
```

## Cause

There are many individual causes, so the 20 most frequent are shown and
the remainder collapsed into "Other".

```{r}
#| label: uni-cause
#| fig-height: 8
mort_count |>
  mutate(cause = fct_lump_n(cause, n = 20)) |>
  ggplot(aes(y = fct_rev(fct_infreq(cause)))) +
  geom_bar(fill = "#3b528b") +
  labs(title = "Deaths by cause (top 20)", x = "Deaths", y = NULL)
```

## Summary table

A Table 1 summarising the categorical variables of interest across all
deaths. Counts and column percentages are shown for each level.

```{r}
#| label: uni-table1
mort_count |>
  select(sesso, asst, group, cause, mechanism) |>
  mutate(cause = fct_lump_n(cause, 15)) |>
  tbl_summary(
    label = list(
      sesso     ~ "Sex",
      asst ~ "ASST (2022-2024)",
      group     ~ "Cause group",
      cause     ~ "Cause",
      mechanism ~ "Mechanism"
    )
  ) |>
  bold_labels()
```

# Bivariate distributions

## Cause-group composition within each ASST

Each bar is filled to 100%, so it shows the *share* of deaths
attributable to each cause group within an ASST, making the mix
comparable across areas of different size.

```{r}
#| label: bi-asst-group
#| fig-height: 7
ggplot(mort_count, aes(y = fct_rev(fct_infreq(asst)), fill = group)) +
  geom_bar(position = "fill") +
  scale_x_continuous(labels = scales::percent) +
  pal_fill() +
  labs(title = "Cause-group composition within each ASST",
       x = "Share of deaths", y = NULL, fill = "Cause group")
```

## Age pattern of mortality by cause group

Deaths are aggregated to a count per single year of age within each
cause group; each point is one age × group cell (`alpha = 0.2` to cope
with overplotting), and the smoothed line summarises how mortality
shifts across the age range — different diseases concentrate at clearly
different ages.

```{r}
#| label: bi-group-age
#| fig-height: 6
age_group <- mort_count |>
  count(group, eta, name = "deaths")

ggplot(age_group, aes(x = eta, y = deaths, colour = group)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE, linewidth = 0.8) +
  facet_wrap(~ group, scales = "free_y") +
  pal_col() +
  labs(title = "Age pattern of mortality by cause group",
       x = "Age (years)", y = "Deaths", colour = "Cause group")
```

## Mechanism by cause group

How the avoidability mechanisms map onto cause groups. Each bar is a
cause group, stacked by the count of deaths in each mechanism, so the
plot shows both which groups carry the most deaths and the mix of
mechanisms within each. (A given cause group can span more than one
mechanism — e.g. the 50/50 split causes.)

```{r}
#| label: bi-group-mechanism
#| fig-height: 6
ggplot(mort_count, aes(y = fct_rev(fct_infreq(group)), fill = mechanism)) +
  geom_bar() +
  pal_fill() +
  labs(title = "Deaths by cause group and mechanism",
       x = "Deaths", y = NULL, fill = "Mechanism")
```

::: {.callout-tip}
To compare the mechanism *mix* across groups regardless of size, switch
to `geom_bar(position = "fill")` and a percent x-axis; to read it the
other way round — which groups feed each mechanism — swap the `y` and
`fill` aesthetics.
:::
