---
title: "Stroke sub-model"
date: today
format:
  html: 
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Sections"
    embed-resources: true
    fig-width: 7.5
    fig-height: 6
    fig-align: center
    tbl-cap-location: top
    df-print: kable
    code-tools: false
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "png"
    dpi: 300
    fig.path: "../output/figures/stroke-"
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(targets)

tar_source(here::here("R"))
source(here::here("reports", "_thesis_setup.R"))

tar_load(c(
  area_shp, stroke_centres, stroke_times, stroke_area, diag_stroke,
  smr_geo_full, smr_geo_tracer, smr_geo_allage, C_matrix,
  control_fits, control_table, control_diagnostics,
  stroke_times_table, exposure_contrast_table, tracer_per_10min,
  tracer_engine_table, tracer_exposure_table,
  outcome_feasibility_table, allage_table, allage_diagnostics,
  mort_raw
))
```

::: callout-note
## Could this design have detected an effect if one existed?
A null coefficient from an exposure with no contrast, or from
an outcome the exposure cannot plausibly move, is not evidence of no
effect, and the two are indistinguishable in a coefficient table.
:::

# 1. The stroke network

## 1.1 Centres

```{r centre-table}
stroke_centre_table(stroke_centres) |>
  kable(col.names = c("ID", "Facility", "Level", "Ente", "Comune", "Prov.",
                      "Lat", "Lon", "Geocoded via", "Map"),
        caption = "Every accredited stroke centre used in the routing, including those outside the ATS Milano territory.")
```

```{r centre-source-summary}
sf::st_drop_geometry(stroke_centres) |>
  count(is_hub, geocode_source) |>
  mutate(is_hub = ifelse(is_hub, "Hub (SU II)", "Spoke (SU I)")) |>
  kable(col.names = c("Level", "Geocoding source", "n"),
        caption = "How each centre was located. Manual overrides and Nominatim results deserve the closest look.")
```


```{r fig-centres, fig.cap="Accredited stroke centres over the study area. Centres outside the ATS Milano boundary are included in the routing because residents near the edge may be closer to them."}
plot_stroke_centres(
  area_shp, stroke_centres,
  caption = "Triangles: level-II hubs performing mechanical thrombectomy. Circles: spoke stroke units."
)
```

# 1.2 Travel time

```{r fig-hub-time, fig.cap="Population-weighted travel time to the nearest level-II thrombectomy hub."}
plot_travel_time(smr_geo_full, "t_hub_mean", centres = stroke_centres,
                 caption = "Routed on the OSM network with AREU emergency speeds, from population-weighted census-section origins.")
```

```{r fig-centre-time, fig.cap="Population-weighted travel time to the nearest accredited stroke centre of any level."}
plot_travel_time(smr_geo_full, "t_centre_mean", centres = NULL,
                 caption = "Includes spoke units, which give thrombolysis but not thrombectomy.")
```

```{r time-table}
stroke_times_table |>
  select(-column) |>
  mutate(across(where(is.numeric), \(x) fmt_num(x, 1))) |>
  kable(col.names = c("Measure", "n", "Missing", "Mean", "SD", "Min", "P25",
                      "Median", "P75", "Max"),
        caption = "Travel-time distribution across the 279 areal units.")
```

```{r routing-diag}
tibble::tibble(
    Diagnostic = c("Population covered by the routing",
                 "Mean time to nearest centre (min)",
                 "Mean time to nearest hub (min)",
                 "Residents >45 min from a hub",
                 "Largest centroid-vs-section bias (min)"),
  Value = c(fmt_n(diag_stroke$headline$pop_covered),
            fmt_num(diag_stroke$headline$mean_to_centre, 1),
            fmt_num(diag_stroke$headline$mean_to_hub, 1),
            fmt_pct(100 * diag_stroke$headline$share_over_45),
            fmt_num(diag_stroke$headline$max_centroid_bias, 1))
  ) |>
  kable(caption = "The last row is what routing from area centroids would have cost: the largest per-area discrepancy against the population-weighted section mean.")
```


# 2. Outcome definition

## 2.1 Original proposition (Cerebrovascular mortality, ages 0–74)

Cerebrovascular mortality, ages 0–74, 2022–2024, indirectly standardised
for age and sex. This is the tracer outcome.

```{r fig-cvd-map, fig.cap="Cerebrovascular SMR by areal unit. Hubs overlaid."}
g <- sf::st_as_sf(smr_geo_tracer)
g$cvd_smr <- g$cvd_obs / g$cvd_exp

ggplot(g) +
  geom_sf(aes(fill = cvd_smr), colour = NA) +
  geom_sf(data = sf::st_transform(sf::st_as_sf(stroke_centres)[stroke_centres$is_hub, ],
                                  sf::st_crs(g)),
          shape = 24, fill = "#d94801", colour = "grey10", size = 2.2) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#8c2d04",
                       midpoint = 1, name = "SMR") +
  theme_atlas(map = TRUE)
```

Thrombectomy acts on cerebral infarction (I63) alone; neurosurgical and neurocritical care act on the haemorrhagic subtypes
(I60-I62).

```{r subtype}
sub <- mort_raw |>
  mutate(key = toupper(gsub("[^A-Za-z0-9]", "", causa))) |>
  filter(startsWith(key, "I6"), eta < 75) |>
  mutate(k3 = substr(key, 1, 3),
         subtype = case_when(
           k3 == "I63"                ~ "I63 Cerebral infarction (thrombectomy)",
           k3 %in% c("I60","I61","I62") ~ "I60-I62 Haemorrhagic (neurosurgery)",
           k3 == "I64"                ~ "I64 Stroke, type not specified",
           k3 == "I69"                ~ "I69 Sequelae",
           TRUE                       ~ paste(k3, "Other cerebrovascular")
         )) |>
  count(subtype, name = "deaths") |>
  mutate(pct = 100 * deaths / sum(deaths)) |>
  arrange(desc(deaths))

sub |>
  mutate(pct = fmt_pct(pct)) |>
  kable(col.names = c("Subtype", "Deaths", "%"),
        caption = "Cerebrovascular deaths under 75, by subtype. Note which channel each belongs to.")
```

```{r dilution, results="asis"}
i63  <- sum(sub$deaths[grepl("^I63", sub$subtype)])
haem <- sum(sub$deaths[grepl("^I60", sub$subtype)])
tot  <- sum(sub$deaths)
lvo <- 0.20; win <- 0.50
thromb <- (i63 / tot) * lvo * win

cat(sprintf(paste0(
  "\n> **Dilution.** If the exposure acted through thrombectomy alone it could ",
  "touch %s of this outcome (%s I63 x ~%.0f%% large-vessel occlusion x ~%.0f%% ",
  "presenting in window). A 30%% mortality reduction in that subset would move ",
  "total cerebrovascular mortality by %s - far below anything this design can ",
  "resolve. Adding the neurosurgical channel (%s haemorrhagic) raises the ",
  "reachable fraction to about %s, which is the same order as the credible ",
  "interval in §5.5. **This is the argument for reframing the exposure rather ",
  "than narrowing the outcome.**\n\n"),
  fmt_pct(100*thromb), fmt_pct(100*i63/tot), 100*lvo, 100*win,
  fmt_pct(100*thromb*0.30), fmt_pct(100*haem/tot),
  fmt_pct(100*(thromb + haem/tot))))
```

```{r cvd-counts}
tibble::tibble(
  Quantity = c("Cerebrovascular deaths, 0-74, 2022-2024",
               "Deaths per year",
               "Areal units with zero deaths",
               "Median expected count per area",
               "Areal units with expected count < 5"),
  Value = c(
    fmt_n(sum(g$cvd_obs)),
    fmt_n(sum(g$cvd_obs) / 3),
    fmt_n(sum(g$cvd_obs == 0)),
    fmt_num(median(g$cvd_exp), 1),
    fmt_n(sum(g$cvd_exp < 5))
  )
) |>
  kable(caption = "The tracer outcome. Sparse counts widen every interval that follows.")
```

```{r fig-raw, fig.cap="Cerebrovascular SMR against travel time to the nearest hub. Point size is the expected count, so the large points are the areas the model actually learns from."}
plot_tracer_raw(smr_geo_tracer, "t_hub_mean")
```

```{r layer-variance}
ec <- exposure_contrast(smr_geo_full, C = C_matrix)
hub <- ec[ec$measure == "t_hub_mean", ]

cat(sprintf(
  paste0("> The exposure has a Moran's I of %s, i.e. it is a smooth spatial ",
         "surface. It therefore competes directly with the BYM2 random ",
         "effect for the same variation — see §5.2.\n\n"),
  fmt_num(hub$moran_i, 3))
  )

```

Thus, the original proposition could never show an effect.

## 2.2 Model

A level-II centre is not only a thrombectomy centre. It provides
neurosurgical and neurocritical care, and that channel acts on
haemorrhagic stroke (which is the majority of premature
cerebrovascular deaths here).

Organised stroke-unit care of any level also has better evidence for a
population mortality effect than thrombectomy specifically, so both
destination sets are fitted.

```{r exposure-table}
tracer_exposure_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Destination set", "Term", "RR per SD", "2.5%", "97.5%",
                      "ESS", "R-hat"),
        caption = "Cerebrovascular mortality 0-74 against travel time to two different destination sets.")
```

The exposure is a smooth spatial surface and so is the BYM2 random
effect. When two terms describe the same variation, the data cannot say
which owns it and the covariate coefficient is pulled toward zero
regardless of whether an effect exists.

**Watch the effective sample size**, not just the interval. A low ESS
concentrated on the exposure coefficient, while the intercept samples
freely, is the signature of a covariate trading off against the random
effect.

```{r engine-table}
tracer_engine_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Specification", "Term", "RR per SD", "2.5%", "97.5%",
                      "ESS", "R-hat"),
        caption = "The same tracer under three treatments of spatial structure.")
```


## 2.3 Control panel

```{r control-expectations}
control_table |>
  select(role, expectation) |>
  kable(col.names = c("Role", "Pre-specified expectation"),
        caption = "Written before the estimates were seen.")
```

```{r control-table}
control_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(role, outcome, label, estimate, ci_low, ci_high, verdict) |>
  kable(col.names = c("Role", "Outcome", "Exposure", "RR per SD", "2.5%",
                      "97.5%", "Verdict"),
        caption = "Tracer and controls. The verdict column checks direction against the pre-specified expectation, not merely whether the interval clears the null.")
```

# 3. Ischemi all-age outcomes

The 75-year ceiling belongs to the OECD/Eurostat avoidability
definition, not to the clinical question. An ischaemic stroke at 90 is
as treatable as one at 60.

```{r feasibility}
outcome_feasibility_table |>
  mutate(
    outcome = recode(outcome,
                     i63    = "I63 cerebral infarction, all ages",
                     cvdall = "I60-I69 all cerebrovascular, all ages",
                     haem   = "I60-I62 haemorrhagic, under 75"),
    across(c(mean_per_area, median_per_area, median_expected), \(x) fmt_num(x, 1)),
    across(c(pct_zero, pct_expected_under5), \(x) fmt_pct(x))
  ) |>
  select(outcome, deaths, mean_per_area, median_per_area, max_per_area,
         n_zero, pct_zero, median_expected, verdict) |>
  kable(col.names = c("Outcome", "Deaths", "Mean/area", "Median/area",
                      "Max/area", "Areas with 0", "% zero", "Median expected",
                      "Verdict"),
        caption = "Feasibility of each candidate outcome at 279-area resolution. Compare against the 0-74 avoidable cerebrovascular outcome, which has 851 deaths.")
```

## 3.1 Model

```{r allage-table}
allage_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Outcome ~ exposure", "Term", "RR per SD", "2.5%",
                      "97.5%", "ESS", "R-hat"),
        caption = "All-age outcomes. Read alongside the feasibility verdict above: a tight interval on an unmodellable outcome is the prior speaking.")
```

```{r allage-diag}
allage_diagnostics |>
  mutate(across(c(rhat_max, rho), \(x) fmt_num(x, 3)),
         pct_divergent = fmt_pct(pct_divergent)) |>
  select(model, rhat_max, ess_min, n_divergent, pct_divergent, rho) |>
  kable(col.names = c("Model", "Max R-hat", "Min ESS", "Divergent",
                      "% divergent", "rho"),
        caption = "Convergence for the all-age models.")
```

## 3.2 Maps

```{r fig-i63-map, fig.cap="I63 (cerebral infarction) SMR, all ages, with thrombectomy hubs overlaid."}
gi <- sf::st_as_sf(smr_geo_allage)
ggplot(gi) +
  geom_sf(aes(fill = i63_smr), colour = NA) +
  geom_sf(data = sf::st_transform(
            sf::st_as_sf(stroke_centres)[stroke_centres$is_hub, ], sf::st_crs(gi)),
          shape = 24, fill = "#d94801", colour = "grey10", size = 2.2) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#8c2d04",
                       midpoint = 1, na.value = "grey88", name = "SMR") +
  theme_atlas(map = TRUE)
```

```{r fig-i63-raw, fig.cap="All-age I63 SMR against travel time to the nearest hub."}
d <- sf::st_drop_geometry(smr_geo_allage)
ggplot(d, aes(x = t_hub_mean, y = i63_smr)) +
  geom_hline(yintercept = 1, linetype = "22", colour = "grey45") +
  geom_point(aes(size = i63_exp), alpha = 0.5, colour = "#8c2d04") +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
              colour = "grey25", linewidth = 0.6) +
  scale_size_area(name = "Expected deaths", max_size = 5) +
  labs(x = "Population-weighted travel time to hub (minutes)",
       y = "I63 SMR, all ages",
       subtitle = sprintf("Spearman = %.3f",
                          cor(d$t_hub_mean, d$i63_smr, method = "spearman",
                              use = "complete.obs"))) +
  theme_atlas()
```

# 4. Interpretation

A null tracer coefficient is consistent with at least four different
states of the world, and this design cannot fully separate them. State
which one you believe, and why, rather than reporting "no association".

1.  **No effect.** Access to thrombectomy does not affect
    population-level cerebrovascular mortality in this territory.
2.  **No contrast.** There is an effect, but every area is close enough
    that the study contains no exposure gradient to detect it with. §4
    answers this directly.
3.  **Wrong outcome.** Mortality is incidence times case fatality.
    Access can only move case fatality, and only for the
    thrombectomy-eligible subset — ischaemic strokes, presenting in
    window, large-vessel occlusion. Deaths from haemorrhagic stroke,
    deaths before contact with services, and long-run sequelae all sit
    in the numerator and dilute the signal toward zero.
4.  **Not identified.** The exposure and the spatial random effect
    describe the same smooth variation. §5.2 addresses this.

```{r nco-reading, results="asis"}
tr  <- control_table[control_table$role == "Tracer", ]
nco <- control_table[control_table$role == "Negative control outcome", ]

if (nrow(tr) && nrow(nco)) {
  cat(sprintf(
    paste0("\2n**The negative control.** Travel time is associated with ",
           "all-cancer mortality at RR %s (%s to %s), against %s (%s to %s) ",
           "for the tracer. %s\n\n"),
    fmt_num(nco$estimate, 3), fmt_num(nco$ci_low, 3), fmt_num(nco$ci_high, 3),
    fmt_num(tr$estimate, 3), fmt_num(tr$ci_low, 3), fmt_num(tr$ci_high, 3),
    if (abs(log(nco$estimate)) > abs(log(tr$estimate)))
      paste0("The negative control association is **larger** than the tracer ",
             "association. Whatever travel time is measuring here, it is not ",
             "specific to time-critical care: it tracks something about area ",
             "character that also predicts cancer mortality. The tracer null ",
             "cannot be read as evidence about the stroke pathway until this ",
             "is explained.")
    else
      paste0("The negative control is smaller than the tracer association, ",
             "which is the pattern the design was built to detect.")
  ))
}
```
