---
title: "Stroke sub-model"
date: today
format:
  html: 
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Sections"
    embed-resources: true
    fig-width: 7.5
    fig-height: 6
    fig-align: center
    tbl-cap-location: top
    df-print: kable
    code-tools: false
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "png"
    dpi: 300
    fig.retina: 1
    fig.path: "../output/figures/stroke-"
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(targets)

tar_source(here::here("R"))
source(here::here("reports", "_thesis_setup.R"))

tar_load(c(
  area_shp, stroke_centres, stroke_times, stroke_area, diag_stroke,
  smr_geo_full, smr_geo_tracer, smr_geo_allage, C_matrix,
  control_fits, control_table, control_diagnostics,
  aug_cvd, aug_i63, aug_haem, diag_cvd_bym2, diag_i63_bym2, fig_haem_pair,
  stroke_times_table, exposure_contrast_table, tracer_per_10min,
  tracer_engine_table, tracer_exposure_table, i63_engine_table,
  outcome_feasibility_table, allage_table, allage_diagnostics,
  stroke_subtypes, stroke_subtypes_under75,
  i63_smoothing, i63_exceedance,
  mort_raw
))
```

# 1. The stroke network

## 1.1 Centres

```{r centre-table}
stroke_centre_table(stroke_centres) |>
  kable(col.names = c("ID", "Facility", "Level", "Ente", "Comune", "Prov.",
                      "Lat", "Lon", "Geocoded via", "Map"),
        caption = "Every accredited stroke centre used in the routing, including those outside the ATS Milano territory.")
```

```{r fig-centres, fig.cap="Accredited stroke centres over the study area. Centres outside the ATS Milano boundary are included in the routing because residents near the edge may be closer to them."}
plot_stroke_centres(
  area_shp, stroke_centres,
  caption = "Triangles: level-II hubs performing mechanical thrombectomy. Circles: spoke stroke units."
)
```

## 1.2 Travel time

```{r fig-hub-time, fig.cap="Population-weighted travel time to the nearest level-II thrombectomy hub."}
plot_travel_time(smr_geo_full, "t_hub_mean", centres = stroke_centres,
                 caption = "Routed on the OSM network with AREU emergency speeds, from population-weighted census-section origins.")
```

```{r fig-centre-time, fig.cap="Population-weighted travel time to the nearest accredited stroke centre of any level."}
plot_travel_time(smr_geo_full, "t_centre_mean", centres = NULL,
                 caption = "Includes spoke units, which give thrombolysis but not thrombectomy.")
```

```{r time-table}
stroke_times_table |>
  select(-column) |>
  mutate(across(where(is.numeric), \(x) fmt_num(x, 1))) |>
  kable(col.names = c("Measure", "n", "Missing", "Mean", "SD", "Min", "P25",
                      "Median", "P75", "Max"),
        caption = "Travel-time distribution across the 279 areal units.")
```

```{r routing-diag}
tibble::tibble(
    Diagnostic = c("Population covered by the routing",
                 "Mean time to nearest centre (min)",
                 "Mean time to nearest hub (min)",
                 "Residents >45 min from a hub",
                 "Largest centroid-vs-section bias (min)"),
  Value = c(fmt_n(diag_stroke$headline$pop_covered),
            fmt_num(diag_stroke$headline$mean_to_centre, 1),
            fmt_num(diag_stroke$headline$mean_to_hub, 1),
            fmt_pct(100 * diag_stroke$headline$share_over_45),
            fmt_num(diag_stroke$headline$max_centroid_bias, 1))
  ) |>
  kable(caption = "The last row is what routing from area centroids would have cost: the largest per-area discrepancy against the population-weighted section mean.")
```

# 2. Outcome definition

## 2.1 Original proposition (Cerebrovascular mortality, ages 0–74)

Cerebrovascular mortality, ages 0–74, 2022–2024, indirectly standardised
for age and sex. This is the original tracer outcome.

```{r fig-cvd-map, fig.cap="Cerebrovascular SMR by areal unit, raw (unsmoothed). Hubs overlaid.", fig.width=7.5, fig.height=5.5}
g <- sf::st_as_sf(smr_geo_tracer)
g$cvd_smr <- g$cvd_obs / g$cvd_exp

ggplot(g) +
  geom_sf(aes(fill = cvd_smr), colour = NA) +
  geom_sf(data = sf::st_transform(sf::st_as_sf(stroke_centres)[stroke_centres$is_hub, ],
                                  sf::st_crs(g)),
          shape = 24, fill = "#d94801", colour = "grey10", size = 2.2) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#8c2d04",
                       midpoint = 1, name = "SMR") +
  theme_atlas(map = TRUE)
```

The map above is the raw ratio. With a median of
`r fmt_num(median(g$cvd_exp), 1)` expected deaths per area, a single extra
death moves a small NIL's ratio by a large fraction, so much of that structure
is noise.

## 2.2 Model

Four specifications are fitted to every outcome in this report. They are not
competing estimates of one quantity: each answers a different question, and a
null means something different in each.

| Specification | The question it answers | What a null means here |
|---|---|---|
| **GLM + travel time** | Across the whole territory, do areas further from a hub die more? | No marginal association. Confounded by everything that varies spatially — deprivation, urbanicity, age structure beyond the offset. |
| **BYM2, intercept only** | Where is risk, once small-number noise is shrunk away? | Descriptive; there is no null. Its `rho` says how much residual variation is spatially structured at all. |
| **BYM2 + travel time** | *Within a neighbourhood of similar spatial risk*, does travel time add anything? | Either no effect, or the exposure is not separable from the random effect. Travel time is itself a smooth spatial surface, so these two compete for the same variation. |
| **ESF + travel time** | The same question, partitioning the spatial variation differently. | Agreement with BYM2 means the answer does not hinge on how the spatial term is parameterised. Disagreement is itself the finding. |


```{r exposure-table}
tracer_exposure_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Destination set", "Term", "RR per SD", "2.5%", "97.5%",
                      "ESS", "R-hat"),
        caption = "Cerebrovascular mortality 0-74 against travel time to two different destination sets.")
```

The exposure is a smooth spatial surface and so is the BYM2 random
effect. When two terms describe the same variation, the data cannot say
which owns it and the covariate coefficient is pulled toward zero
regardless of whether an effect exists.

A low Effective sample size (ESS) concentrated on the exposure coefficient, 
while the intercept samples freely, is the signature of a covariate trading
off against the random effect.

```{r engine-table}
tracer_engine_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Specification", "Term", "RR per SD", "2.5%", "97.5%",
                      "ESS", "R-hat"),
        caption = "The same tracer under three treatments of spatial structure.")
```

```{r fig-cvd-pair, fig.cap="Cerebrovascular mortality 0\u201374, raw against BYM2-smoothed, on a shared scale.", fig.width=11, fig.height=5.2}
p_raw <- plot_smr_map(g, value = "cvd_smr", title = NULL, subtitle = NULL,
                      caption = NULL) |> overlay_hubs(stroke_centres)
p_bym <- plot_smr_map(aug_cvd, value = "bym2_rr", title = NULL, subtitle = NULL,
                      caption = NULL) |> overlay_hubs(stroke_centres)

map_pair(p_raw, p_bym,
         titles = c("Raw SMR (observed / expected)",
                    "BYM2-smoothed relative risk"))
```

The smoothing model is intercept-only and contains no travel time. Adjusting
the descriptive surface for the exposure under test would remove from the map
whatever signal the analysis is looking for.

```{r cvd-smoothing-diag}
tibble::tibble(
  Quantity = c(
    "Mixing parameter rho (share of residual variation that is spatially structured)",
    "Range of raw SMR", "Range of smoothed RR",
    "Max R-hat", "Min effective sample size"),
  Value = c(
    fmt_num(diag_cvd_bym2$rho_mean, 3),
    sprintf("%s to %s", fmt_num(min(g$cvd_smr, na.rm = TRUE), 2),
            fmt_num(max(g$cvd_smr, na.rm = TRUE), 2)),
    sprintf("%s to %s", fmt_num(min(aug_cvd$bym2_rr, na.rm = TRUE), 2),
            fmt_num(max(aug_cvd$bym2_rr, na.rm = TRUE), 2)),
    fmt_num(diag_cvd_bym2$rhat_max, 4),
    fmt_n(diag_cvd_bym2$neff_min))
) |>
  kable(caption = "How much the smoothing did. A smoothed range much narrower than the raw range means most of the raw variation was noise; a low rho means little of what remains is spatially structured.")
```

```{r fig-cvd-exceed, fig.cap="Exceedance probability, cerebrovascular mortality 0-74.", fig.width=7.5, fig.height=5.5}
plot_exceedance_map(aug_cvd, title = NULL, subtitle = NULL) |>
  overlay_hubs(stroke_centres)
```

```{r cvd-exceed-count, results="asis"}
exc <- sf::st_drop_geometry(aug_cvd)$bym2_exceed
cat(sprintf(paste0(
  "\n> %s of %s areal units carry at least an 80%% posterior probability that ",
  "cerebrovascular mortality runs more than 20%% above expectation, and %s reach ",
  "95%%.\n\n"),
  fmt_n(sum(exc > 0.80, na.rm = TRUE)), fmt_n(length(exc)),
  fmt_n(sum(exc > 0.95, na.rm = TRUE))))
```

Thrombectomy acts on cerebral infarction (I63) alone; neurosurgical and neurocritical care act on the haemorrhagic subtypes
(I60-I62).

```{r subtype}
stroke_subtypes_under75 |>
  mutate(pct = fmt_pct(pct)) |>
  kable(col.names = c("Subtype", "Deaths", "%"),
        caption = "Cerebrovascular deaths under 75, by subtype. Note which channel each belongs to.")
```

```{r dilution, results="asis"}
cat("\n> **Dilution.** ",
    stroke_subtype_sentence(stroke_subtypes_under75), "\n\n", sep = "")
```

```{r fig-raw, fig.cap="Cerebrovascular SMR against travel time to the nearest hub. Point size is the expected count, so the large points are the areas the model actually learns from.", fig.width=7.5, fig.height=4.5}
plot_tracer_raw(smr_geo_tracer, "t_hub_mean")
```

```{r layer-variance}
ec <- exposure_contrast(smr_geo_full, C = C_matrix)
hub <- ec[ec$measure == "t_hub_mean", ]

cat(sprintf(
  paste0("> The exposure has a Moran's I of %s, i.e. it is a smooth spatial\n",
         "surface. It therefore competes directly with the BYM2 random\n",
         "effect for the same variation — see §2.2.\n\n"),
  fmt_num(hub$moran_i, 3))
  )
```

## 2.3 Control panel

```{r control-expectations}
control_table |>
  select(role, expectation) |>
  kable(col.names = c("Role", "Pre-specified expectation"),
        caption = "Written before the estimates were seen.")
```

```{r control-table}
control_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(role, outcome, label, estimate, ci_low, ci_high, verdict) |>
  kable(col.names = c("Role", "Outcome", "Exposure", "RR per SD", "2.5%",
                      "97.5%", "Verdict"),
        caption = "Tracer and controls. The tracer arm is fitted on the reference outcome (I63, all ages), not on the 0-74 cerebrovascular outcome of §2. The verdict column checks direction against the pre-specified expectation, not merely whether the interval clears the null.")
```

Thus, the original proposition could never show an effect.

# 3. Ischemic all-age outcomes

The 75-year ceiling belongs to the OECD/Eurostat avoidability
definition, not to the clinical question. An ischaemic stroke at 90 is
as treatable as one at 60.

```{r feasibility}
outcome_feasibility_table |>
  mutate(
    outcome = recode(outcome,
                     i63    = "I63 cerebral infarction, all ages",
                     cvdall = "I60-I69 all cerebrovascular, all ages",
                     haem   = "I60-I62 haemorrhagic, under 75"),
    across(c(mean_per_area, median_per_area, median_expected), \(x) fmt_num(x, 1)),
    across(c(pct_zero, pct_expected_under5), \(x) fmt_pct(x))
  ) |>
  select(outcome, deaths, mean_per_area, median_per_area, max_per_area,
         n_zero, pct_zero, median_expected, verdict) |>
  kable(col.names = c("Outcome", "Deaths", "Mean/area", "Median/area",
                      "Max/area", "Areas with 0", "% zero", "Median expected",
                      "Verdict"),
        caption = "Feasibility of each candidate outcome at 279-area resolution. Compare against the 0-74 avoidable cerebrovascular outcome, which has 851 deaths.")
```

## 3.1 Does travel time predict risk?


```{r i63-engine-table}
i63_engine_table |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Specification", "Term", "RR per SD", "2.5%", "97.5%",
                      "ESS", "R-hat"),
        caption = "I63, all ages, against travel time to the nearest thrombectomy hub. Compare row for row with the cerebrovascular 0-74 table in §2.2.")
```

```{r i63-engine-verdict, results="asis"}
te <- i63_engine_table
g_est <- te$estimate[grepl("GLM", te$model)]
b_est <- te$estimate[te$model == "BYM2"]

if (length(g_est) && length(b_est)) {
  cat(if (abs(log(g_est)) > 2 * abs(log(b_est)))
    paste0("\n> The association is substantially larger without the spatial ",
           "term. That is an identification result rather than a substantive ",
           "one: the BYM2 random effect is absorbing the travel-time gradient, ",
           "and neither estimate should be read as the effect of access.\n\n")
  else if (abs(log(b_est)) < 0.02 && abs(log(g_est)) < 0.02)
    paste0("\n> Both specifications are null. Removing the age ceiling has not ",
           "produced an association that the 0\u201374 outcome was hiding, so ",
           "the age restriction was not what was suppressing the signal.\n\n")
  else
    paste0("\n> The estimate is stable across specifications, so it is not an ",
           "artefact of the spatial term competing with the exposure.\n\n"))
}
```

```{r allage-table}
allage_table |>
  filter(!grepl("centre", model)) |>
  mutate(across(c(estimate, ci_low, ci_high), \(x) fmt_num(x, 3))) |>
  select(model, label, estimate, ci_low, ci_high, ess, rhat) |>
  kable(col.names = c("Outcome ~ exposure", "Term", "RR per SD", "2.5%",
                      "97.5%", "ESS", "R-hat"),
        caption = "All-age outcomes under BYM2. Read alongside the feasibility verdict above: a tight interval on an unmodellable outcome is the prior speaking.")
```

```{r allage-diag}
allage_diagnostics |>
  filter(!grepl("centre", model)) |>
  mutate(across(c(rhat_max, rho), \(x) fmt_num(x, 3)),
         pct_divergent = fmt_pct(pct_divergent)) |>
  select(model, rhat_max, ess_min, n_divergent, pct_divergent, rho) |>
  kable(col.names = c("Model", "Max R-hat", "Min ESS", "Divergent",
                      "% divergent", "rho"),
        caption = "Convergence for the all-age models.")
```

## 3.2 Where is the risk? Descriptive surfaces

```{r fig-i63-map, fig.cap="I63 (cerebral infarction) SMR, all ages, raw (unsmoothed), with thrombectomy hubs overlaid.", fig.width=7.5, fig.height=5.5}
gi <- sf::st_as_sf(smr_geo_allage)
ggplot(gi) +
  geom_sf(aes(fill = i63_smr), colour = NA) +
  geom_sf(data = sf::st_transform(
            sf::st_as_sf(stroke_centres)[stroke_centres$is_hub, ], sf::st_crs(gi)),
          shape = 24, fill = "#d94801", colour = "grey10", size = 2.2) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#8c2d04",
                       midpoint = 1, na.value = "grey88", name = "SMR") +
  theme_atlas(map = TRUE)
```

```{r fig-i63-pair, fig.cap="I63 cerebral infarction, all ages: raw against BYM2-smoothed, shared scale.", fig.width=11, fig.height=5.2}
p_raw_i63 <- plot_smr_map(gi, value = "i63_smr", title = NULL, subtitle = NULL,
                          caption = NULL) |> overlay_hubs(stroke_centres)
p_bym_i63 <- plot_smr_map(aug_i63, value = "bym2_rr", title = NULL,
                          subtitle = NULL, caption = NULL) |>
  overlay_hubs(stroke_centres)

map_pair(p_raw_i63, p_bym_i63,
         titles = c("Raw SMR (observed / expected)",
                    "BYM2-smoothed relative risk"))
```

Posterior probability that all-age I63 relative risk exceeds 1.20.

```{r fig-i63-exceed, fig.cap="Exceedance probability, I63 all ages.", fig.width=7.5, fig.height=5.5}
plot_exceedance_map(aug_i63, title = NULL, subtitle = NULL) |>
  overlay_hubs(stroke_centres)
```

### Haemorrhagic stroke, the other channel


```{r fig-haem, fig.cap="Haemorrhagic stroke (I60-I62) under 75: smoothed relative risk and exceedance probability. Not a negative control for hub access \u2014 level-II centres perform neurosurgery as well as thrombectomy \u2014 so this is descriptive only.", fig.width=11, fig.height=5.2}
fig_haem_pair
```

```{r fig-i63-raw, fig.cap="All-age I63 SMR against travel time to the nearest hub.", fig.width=7.5, fig.height=4.5}
d <- sf::st_drop_geometry(smr_geo_allage)
ggplot(d, aes(x = t_hub_mean, y = i63_smr)) +
  geom_hline(yintercept = 1, linetype = "22", colour = "grey45") +
  geom_point(aes(size = i63_exp), alpha = 0.5, colour = "#8c2d04") +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
              colour = "grey25", linewidth = 0.6) +
  scale_size_area(name = "Expected deaths", max_size = 5) +
  labs(x = "Population-weighted travel time to hub (minutes)",
       y = "I63 SMR, all ages",
       subtitle = sprintf("Spearman = %.3f",
                          cor(d$t_hub_mean, d$i63_smr, method = "spearman",
                              use = "complete.obs"))) +
  theme_atlas()
```

# 4. Interpretation

A null tracer coefficient is consistent with at least four different
states of the world.

1.  **No effect.** Access to thrombectomy does not affect
    population-level cerebrovascular mortality in this territory.
2.  **No contrast.** There is an effect, but every area is close enough
    that the study contains no exposure gradient to detect it with. §1.2
    answers this directly.
3.  **Wrong outcome.** Mortality is incidence times case fatality.
    Access can only move case fatality, and only for the
    thrombectomy-eligible subset — ischaemic strokes, presenting in
    window, large-vessel occlusion. Deaths from haemorrhagic stroke,
    deaths before contact with services, and long-run sequelae all sit
    in the numerator and dilute the signal toward zero.
4.  **Not identified.** The exposure and the spatial random effect
    describe the same smooth variation. §2.2 addresses this.
